--- 
title: "Claudia's data"
author: "Philipp C. Muench"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: scrbook
link-citations: yes
colorlinks: yes
graphics: yes
lot: yes
lof: yes
toc-depth: 2
fontsize: 11pt
mainfont: Palatino
monofont: "Source Code Pro"
monofontoptions: "Scale=0.8"
outputbo:
  bookdown::gitbook:
    config:
      sharing: null
      toc:
        collapse: section
      fontsettings:
        theme: white
        family: Source Code Pro
        size: 2
---

```{r setup, include=FALSE}
set.seed(0728)
knitr::opts_chunk$set(tidy = TRUE)
if (knitr::is_html_output()) {
  # ignore percentage widths for HTML output, unless they are used for mutiple
  # figures side by side
  knitr::opts_hooks$set(out.width = function(options) {
    if (options$fig.show != 'hold' && grepl('%$', options$out.width))
      options$out.width = NULL
    options
  })
}
```

# Preface {-}

Processing of NGS data is done using **Snakemake** and not covered by this report yet and will be added at a later stage. In short, preprocessing and analysis (such as `bwa` alignment and variant calling using [Lofreq](https://csb5.github.io/lofreq/commands/) is defined using a `Snakefile` and a set of snakemake *rules* which are run on the HZI server (host) and detailed analysis (such as generation of figures and statistical tests) are done using R markdown on the *client* machine.

## How to generate this report  {-}

In [vscode](https://code.visualstudio.com/) with bookdown addon a server can be started using `serve_book`

```{r, eval=FALSE}
bookdown::serve_book(getwd())
```
This will starts a browser window pointing to [http://127.0.0.1:7853](http://127.0.0.1:7853) presenting the report written in [R Markdown](https://rmarkdown.rstudio.com/)

## Acknowledgements {-}

