--- 
title: "OMM report"
author: "Philipp C. Muench"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: scrbook
link-citations: yes
colorlinks: yes
graphics: yes
lot: yes
lof: yes
toc-depth: 2
fontsize: 11pt
mainfont: Palatino
monofont: "Source Code Pro"
monofontoptions: "Scale=0.8"
css: style.css
outputbo:
  bookdown::gitbook:
    css: style.css
    config:
      sharing: null
      toc:
        collapse: section
      fontsettings:
        theme: white
        family: Source Code Pro
        size: 2
---

```{r setup, include=FALSE}
set.seed(0728)
knitr::opts_chunk$set(tidy = TRUE)
if (knitr::is_html_output()) {
  # ignore percentage widths for HTML output, unless they are used for mutiple
  # figures side by side
  knitr::opts_hooks$set(out.width = function(options) {
    if (options$fig.show != 'hold' && grepl('%$', options$out.width))
      options$out.width = NULL
    options
  })
}
```

# Setup {#Setup}

## Installation

```{r, results="hide"}
rm(list = ls())
source("utils.R")
library("vcfR")
library("rmdformats")
library("ggplot2")
library("reshape2")
library("ggridges")
library("knitr")
library("kableExtra")
library("plotly")
library("DT")
library("circlize")
library("ComplexHeatmap")
library("rtracklayer")
library("bookdown")
```

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```

_Note:_ for `rtracklayer` you might need _curl_
```{bash, eval = F}
sudo apt-get install libcurl4-openssl-dev
```

You need to install [Pandoc](https://pandoc.org/installing.html). 

## Generate report

```{r, eval=FALSE}
setwd("~/Cloud/ActiveProjects/OligoMM-report")
bookdown::serve_book(getwd())
# servr::daemon_stop(1)
```

This will starts a browser window pointing to [http://127.0.0.1:7853](http://127.0.0.1:7853) presenting the report written in [R Markdown](https://rmarkdown.rstudio.com/) and will update as soon as there are changes on `.Rmd` files in the workspace.

To end bookdown server type `servr::daemon_stop(1)` where `1` is the ID of the server running

`bookdown::render_book(getwd())`
## Data structure

`data/raw`
: files from host machine, contains `coverage`, _lofreq_ and _varscan_ output

`data/annotation`
: annotation files for OMM12 genomes e.g. phage location (*pro-hunter*, *fasta* files and *Prokka* annotation)

`data/tables`
: tables in tsv format for e.g. import with Excel

`data/annotation/gff`
: *Prokka* annotation

`data/rda`
: processed raw data in *rds* format generated by `01-load-data.rmd`

`docs`
: `html` files generated via `bookdown::serve_book(getwd())`